---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# xps

<!-- badges: start -->
<!-- badges: end -->

Experience studies are used by actuaries to explore historical experience across
blocks of business and to inform assumption setting activities. This package 
provides tidyverse-insprired functions for preparing data, creating studies, 
and beginning assumption development.

## Installation

You can install the development version of xps from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("mattheaphy/xps")
```

## Simulated data set

The `xps` package includes a data frame containing simulated census data for a theoretical deferred annuity product with an optional guaranteed income rider. The grain of this data is one row per policy.

```{r, warning=FALSE}
library(xps)
library(dplyr)

census_dat
```

The data includes 3 policy statuses: Active, Death, and Surrender.

```{r}
status_counts <- table(census_dat$status)
```

Let's assume we're interested in calculating the probability of surrender over one policy year. We cannot simply calculate the proportion of policies in a surrendered status as this does not represent an annualized surrender rate. 

```{r}
prop.table(status_counts)
```

## Creating exposed data

In order to calculate annual surrender rates, we need to break each policy into multiple records. There should be one row per policy per year.

The `expose_` family of functions is used to perform this transformation. 


```{r example}
exposed_data <- expose(census_dat, end_date = "2019-12-31", 
                        target_status = "Surrender")

exposed_data
```

Now that the data has been "exposed" by policy year, the observed annual surrender probability can be calculated as:

```{r}
sum(exposed_data$status == "Surrender") / sum(exposed_data$exposure)
```

As a default, the `expose` function calculate exposures by policy year. This can also be accomplished with the function `expose_py`. Other implementations of `expose` include:

- `expose_cy` = exposures by calendar year
- `expose_cq` = exposures by calendar quarter
- `expose_pm` = exposures by policy month
- `expose_cm` = exposures by calendar month

## Experience study summary function

The `exp_stats` function can be used to create a summary of observed experience data.

```{r}
exp_stats(exposed_data)
```

If the data frame passed into `exp_stats` is grouped, the resulting output will contain one record for each unique group.

```{r}
library(dplyr)

exp_res <- exposed_data |> 
  group_by(pol_yr, inc_guar) |> 
  exp_stats()

exp_res
```

```{r, echo = FALSE, warning=FALSE, message=FALSE}

library(ggplot2)

.colors <- c("#eb15e4", "#7515eb")

theme_set(theme_light())

exp_res |> 
  ggplot(aes(pol_yr, q_obs, color = inc_guar)) + 
  geom_point() + 
  geom_line() + 
  scale_y_continuous(labels = scales::label_percent(accuracy = 1)) + 
  scale_color_manual(values = .colors) + 
  labs(title = "Observed Surrender Rates by Policy Year and Income Guarantee Presence")
```


To derive actual-to-expected rates, attach one or more columns of expected termination rates to the exposure data. Then, pass these column names to the `expected` argument of `exp_stats`.

```{r}
exposed_data$exp_surr <- ifelse(exposed_data$inc_guar, 0.015, 0.03)

exp_res <- exposed_data |> 
  group_by(pol_yr, inc_guar) |> 
  exp_stats(expected = "exp_surr")

exp_res

```
