---
title: "Other functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Other functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

actxps:::set_actxps_plot_theme()
```


This vignette features functions that are not covered in other vignettes.

```{r packages}
library(actxps)
library(lubridate)
```

## Policy duration functions

The `pol_()` family of functions can be used to calculate policy years, months, quarters, weeks, or any other arbitrary duration. Each function accepts a vector of dates and a vector of issue dates.

**Example**: assume a policy was issued on 2022-05-10 and we are interested in calculating various policy duration values at the end of calendar years 2022-2032.

```{r pol-dur1}


dates <- ymd("2022-12-31") + years(0:10)

# policy years
pol_yr(dates, "2022-05-10")

# policy quarters
pol_qtr(dates, "2022-05-10")

# policy months
pol_mth(dates, "2022-05-10")

# policy weeks
pol_wk(dates, "2022-05-10")


```

The more general `pol_interval()` function can be used to calculate any arbitrary duration. This function has a third argument where the length of the policy duration can be specified. This argument must be a period object. See `lubridate::period()` for more information.

```{r pol-dur2}

# days
pol_interval(dates, "2022-05-10", days(1))

# fortnights
pol_interval(dates, "2022-05-10", weeks(2))

```

## Predictive modeling support functions

The `add_predictions()` function can be used to attach predictions from any model with a `predict()` method.

Below, a very simple logistic regression model is fit to surrender experience in the first ten policy years. Predictions from this model are then added to exposure records using `add_predictions()`. This function only requires a data frame of exposure records and a model with a `predict()` method. Often, it is necessary to specify additional model-specific arguments like `type` to ensure `predict()` returns the desired output. In the example below, `type` is set to "response" to return probabilities instead of the default predictions on the log-odds scale.

The `col_expected` argument is used to rename the column(s) containing predicted values. If no names are specified, the default name is "expected".

```{r add-preds, fig.height=4, fig.width=5}

# create exposure records
exposed_data <- expose(census_dat, end_date = "2019-12-31",
                       target_status = "Surrender") |> 
  filter(pol_yr <= 10) |> 
  # add a response column for surrenders
  mutate(surrendered = status == "Surrender")

# create a simple logistic model
mod <- glm(surrendered ~ pol_yr, data = exposed_data, 
           family = "binomial", weights = exposure)

exp_res <- exposed_data |> 
  # attach predictions
  add_predictions(mod, type = "response", col_expected = "logistic") |> 
  # summarize results
  group_by(pol_yr) |> 
  exp_stats(expected = "logistic")
 
# create a plot
plot_termination_rates(exp_res)

```



In addition, for users of the tidymodels framework, the actxps package includes a recipe step function, `step_expose()`, that can apply the `expose()` function during data pre-processing.

```{r recipe, warning=FALSE}
library(recipes)

recipe(~ ., data = census_dat) |> 
  step_expose(end_date = "2019-12-31", target_status = "Surrender")

```

